### 2014-07-15 paper.tex
updated format

Marek's disease (MD) is an economically significant chicken disease that affects the
poultry industry worldwide with estimated annual loss of
\$2 billion loss has been reported from outbreaks~\cite{morrow2004marek}.
The disease is caused by the highly oncogenic Marek's disease virus (MDV)
an alphaherpesvirus that induces T-cell lymphomas in susceptible birds.
Vaccination is the primary control measure, which is effective in reducing
incidence of tumor formation.
However, since MD vaccines are not sterilizing, they do not prevent infection or
horizontal spread of the virus.
As a consequence, MDV field strains that overcome vaccinal protection have arisen repeatedly over time.
Therefore, there is a need for sustainable alternative controls measures, such as improving genetic
resistance.

Many studies have reported strong associations between MHC alleles and resistance or susceptibility to
MD. For example, chickens with MHC allele B$^{21}$ are highly resistant in contrast to chickens with
B$^{16}$ allele which are highly susceptible.
ADOL lines 6 and 7, both share (B$^6$) allele, yet exhibit different phenotypic responses to
MD; e.g., challenge with JM/102W strain with result in 0 and 100\% MD incidence for lines 6 and 7,
respectively.
Thus, the major unanswered questions are what genetic factors, especially those that are non-MHC,
contribute to susceptibility and resistance of the disease and what is the main contributing
mechanism.

In the past decades, significant efforts have been spent to study variations in a global gene
expression between resistant and susceptible birds using
microarray and RNA-Seq methods in order to identify non-MHC genes that contribute
to resistance to MD~\cite{}.
However, none of the studies have investigated differential expression of alternative isoforms,
which are known to play a significant role in many biological events including
immune responses.
In addition, studies have shown that isoform expression levels can provide better signatures for some
diseases~\cite{zhang2013isoform}.
Changes of isoform expression levels are governed partly by two types of \textit{cis-}regulatory
elements: Exon Splicing Enhancer (ESE) and Exon Splicing Silencer (ESS),
which are located within an exon sequence.
A number of sequence motifs of ESE and ESS have been identified in human and some other organisms
and could be predicted \textit{in silico}.
Mutations that disrupt or create those motifs could alter splicing
patterns leading to aberrant alternative splicing.
A number of disease-associated single-nucleotide polymorphsims in coding regions
(cSNPs) that affect ESEs and ESSs have been well
characterized~\cite{blencowe2000exonic, wang2007splicing}.
%In fact, 15\% of mutations that cause genetic diseases affect pre-mRNA splicing via this mechanism.
Therefore, variations in isoform expression could lead to identification of SNPs
that underlie genetic resistance to MD.
In this article, we reported differential-expressed genes and isoforms that may contribute to
resistance to MD as well as cSNPs that can potentially affect isoform expression levels.

@@

To extensively study gene and isoform expression, we incorporated \textit{Ensembl} gene models
release 73 with \textit{de novo} and a reference-guided transcriptome assembly to build custom
gene models.
The models, therefore, include both \textit{Ensembl} annotated transcripts
and putative transcripts and genes.
The advantage of using custom gene models is it allows an investigation of unannotated genes and
isoforms, which is necessary for in-depth transcriptome study of an immune system.

Many genes were found to be differentially expressed (DE) between control and infected
chickens in lines 6 and line 7.
While the number of unique downregulated genes in both lines was approximately equal,
the number of unique upregulated genes in line 7 was much greater compared
to line 6 (Figure~\ref{degenes_venn})~\cite{lee1983ontogeny}.
Interestingly, some genes that were differentially expressed in both lines were
regulated in the opposite direction (Table~\ref{tab:opposite}).
Among genes downregulated in line 6 but upregulated in line 7 were
\textit{LL} (lung lectin) and \textit{SFTPA1}, which encode a calcium-dependent C-type
lectin and a lung surfactant protein respectively.
Both molecules are important in innate immunity~\cite{hogenkamp2008chicken,kingma2006defense}.
\textit{LIMS1} is involved in cell differentiation and proliferation and
\textit{PPARG} is a suppressor of \textit{NF$\kappa$B}-mediated proinflamatory response.
On the other hand, nearly all genes upregulated in line 6 but downregulated in line 7
are involved in cell survival such as mRNAs splicing, cell growth, and protein synthesis,
except CD7 whose function is involved in T cell-B cell interaction.
This difference suggests that even at this stage of infection in line 6, the lytic phase 
could be subsided, therefore, only genes important for survival of cell are upregulated in line 6.
In comparison, the lytic phase in line 7 may still continue and as a result, genes involved in
immune responses are still upregulated.
In addition, type I interferon (\textit{IFN-$\gamma$} and \textit{IFN-$\beta$}) as well as \textit{INF-$\alpha$3} were found to be highly
upregulated in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors
were not different in line 6, but upregulated in line 7.
This could also reflect the ongoing immune responses in line 7.

Some DE genes have been reported from previous microarray studies were also found to be
differentially expressed in this study.
For example, B6.1 (Bu-1) gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4 d.p.i~\cite{sarson2008transcriptional}.
It was also found to be down-regulated about 3-fold in line 7 (susceptible).
Similarly, \textit{GMZA} gene reported to be upregulated across genetically different chickens
(B$^{19}$, B$^{21}$ alleles), was also found to be highly upregulated.
In contrast, some genes that have been reported to be highly expressed in resistant chickens were
downregulated in both lines. Those genes are \textit{AMIGO2, MMP13} and \textit{CLEC3B}, which were
found to be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed in resistant chickens:
% \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line 7$_{2}$) including
\textit{AVD, ART1, NOS2, CXCL13L2, MX1} and \textit{SOCS1}~\cite{smith2011systems}
were also found to be highly upregulated in both line 6 and 7.
However, our results show the similar expression patterns for \textit{IL6} and \textit{IL18}, which were
only upregulated in line 7 in early stage of infection (3-5 d.p.i).

In contrast, \textit{IL15} has been reported non differential-expressed between control
and infected chickens in
line 6$_{1}$ and 7$_{2}$~\cite{kaiser2003differential}; however,
it was only upregulated in line 7 in our study.
Expression of \textit{IL15} is induced by \textit{TLR9}, which binds to non-methylated CpG
residues present in genomes of many DNA viruses, including herpes simplex virus.
This cytokine auto-regulates the expression of \textit{CD40}, which is
a transmembrane receptor required for activation of macrophages by CD4 T cells.
Consequently, \textit{CD40} was only upregulated in line 7 (data not shown).

@@

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq data~\cite{young2010method}.
Significantly perturbed pathways (FDR $< 0.1$) from both lines that involved in
immune response include TLR signaling pathway, cytokine-cytokine receptor interaction,
intestinal immune network for IgA production, and cell adhesion molecules (CAMs)~\ref{line67_kegg}.
Some other pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling pathway,
NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.
Although MHC I gene (\textit{BF1}) was differentially expressed in both lines, other genes involved in
expressing newly synthesized MHC I were only upregulated in line 7
suggesting that new MHC I molecules were actively produced and expressed on cell surface.
% (the minor class I is upregulated but the major class I is downregulated)
Furthermore, gene ontology analysis of biological processes (GO:BP) shows that categories involved in
both adaptive and innate immune responses were enriched in line 7 (supplementary material).
On the other hand, only categories involved in innate immune responses were enriched in line 6.
In addition, enrichment of apoptosis pathway in line 7 indicates that the programmed
cell death could be induced by CTL response to eliminate the viruses.

@@

Due to those issues, it is not feasible to accurately estimate expression of isoforms, especially when
gene annotation is constructed from \textit{de novo} assembly~\cite{trapnell2013differential}.
To avoid these issues, we chose to study exon expression instead of isoform expression.
Using MISO with exon-centric method, only reads spanning across a few exons are used and only exons
involved in a splicing event are examined.
The expression of exon inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used to
infer the portion of transcripts that include the exon in each sample~\cite{Katz:2010iv}.
In this study, we investigated the three most common alternative splicing events in
vertebrates, which are skipped exons (SE), an alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.

@@

Group I (Table~\ref{tab:line67i_diff_line67u_one}) includes genes with $\Psi$s that were
up- or down-regulated in infected chickens in line 6 only.
This group includes \textit{BCL11B} (B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell lymphoma
associated C2H2-type zinc finger protein encoding gene, which functions as a tumour-suppressor
in T-cell lymphoma in human.
According to homologous alignments on UCSC genome browser, a splice form with the skipped exon
is similar to mouse \textit{BCL11B isoform b}.
The skipped exon was expressed 30\% in infected line 6 chickens; whereas it was rarely expressed
(4-7\%) in the control line 6 and both groups in line 7.
The skipped exon was not found to encode any known protein domain, however, it is in the middle of
two adjacent C2H2-type finger protein domains.
Another gene related to B and T cell lymphomas is \textit{SIK2} (salt-inducible kinase 2).
This gene has been reported to have a negative effect on T cell lymphomas by limiting the transcription of
HTLV-1 virus~\cite{tang2013lkb1}.
Some other genes are important in pre-mRNA splicing including \textit{TRA2A},
\textit{SRSF6} and \textit{GEMIN6}.
\textit{TRA2A} (transformer-2 protein homolog alpha) encodes RNA recognition motif (RRM),
the skipped exon is not found to encode any known protein domain.
\textit{SRSF6} (SR splicing factor 6) encodes a nuclear protein that belongs to the splicing
factor protein family.
\textit{GEMIN6} plays a role in the assembly of spliceosomal snRNP in cytoplasm.
These genes could play a significant role in regulating inclusion of alternative exons.
A gene in this group that might be important for innate immune responses is \textit{RAC3} (Ras-related C3
botulinum toxin substrate 3).
This gene encodes small GTPases, belonging to Ras family, that regulates a wide variety of cellular events
including cell growth, cytoskeletal reorganization, and the activation of protein kinases.
The role of small GTPases in immune responses is discussed further below.

In group II (Table~\ref{tab:line67i_diff_line67u_two}),
 $\Psi$ values were relatively stable in control and infected chickens within line, but not between lines.
Genes that could play an important role in immune responses are \textit{ITGB2} and \textit{HCK}.
\textit{HCK} transmits signal from cell surface receptors such as \textit{FCGR1A, FCGR2A, IL2, IL6,
IL18}, and integrins (\textit{ITGB1, ITGB2}).
\textit{ITGB2} (CD18) encodes subunit $\beta_{2}$ integrin of \textit{LFA-1} and \textit{CR3} receptors.
\textit{LFA-1} plays an important role in adhesion of lymphocytes with other cells.
\textit{CR3} binds to a vast array of ligands and molecules including complement C3bi, microbial proteins,
ICAM-1 and -2, ECM proteins, and coagulation proteins~\cite{}.
It plays a significant role in neutrophils and monocytes activation including phagocytosis, adhesion and
migration~\cite{}.
The role of \textit{ITGB2} in immune responses is discussed further in the next section.
\textit{DYNLT1,DYNLL2, SEPT11, PFN2}, and \textit{ZDHHC7} are also involved in cell
rearrangement and cytokinesis.
In particular, \textit{DYNLT1} and \textit{DYNLL2} are dynein proteins that have been demonstrated to
regulate T cell activation by driving T cell receptor microclusters (TCR-MCs)
toward the center of immune synapse~\cite{hashimoto2011dynein}.

Group III (Table~\ref{tab:line67i_diff_line67u_three})
includes genes that exhibit differential isoform expression only in infected line 7.
A number of genes in this group encode proteins that are parts of
spliceosome: \textit{SRSF3}, \textit{HNRNPDL},
\textit{SFSWAP}, \textit{THOC1}, \textit{RNPC3} and \textit{SRSF5}.
Two genes involved in cell-cell contact regulated by integrins are also in this group.
\textit{PPP1R12A} is a myosin phosphatase that regulates the interaction of actin
and myosin downstream of GTPase Rho proteins.
\textit{PODXL} encodes PODX-like protein that functions in integrin-dependent
manner as both pro-adhesive and anti-adhesive molecules.
This protein is involved in cell-to-cell contact, cell trafficking, and cancer progression~\cite{}.

The last group (Group IV, Table~\ref{tab:line67i_diff_line67u_three})
only has one gene, \textit{GOSR1}.
The $\Psi$ value of this gene was less than 0.20 cutoff in control and infected chickens in line 6, but it is
significantly different between infected chickens in line 6 and 7.
Also, there is a significant difference between control and infected chickens in line 7.
This gene encodes a trafficking membrane protein important for transporting proteins
from \textit{cis-} to \textit{trans-}golgi network.

\subsubsection*{Roles of \textit{LFA-1} and actin cytoskeleton pathway in T cells activation}

Integrins are a special kind of receptors that transmit signals bidirectionally across the cell membrane.
They are heterodimeric composed of an $\alpha$ (large) and a $\beta$ (small)
subunits~\cite{wang2010immunopathologies}.
$\beta_{2}$ (CD18) subunit encoded by \textit{ITGB2} is expressed on lymphocytes and antigen presenting cells
(APCs) as a component of \textit{LFA-1} and \textit{CR3} receptors.
LFA-1 binds to its ligand ICAM-1 to help form a synapse that bring APC and T cell together to
initiate antigen presentation leading to T cell activation~\cite{dustin2000immunological}.
Absence of LFA-1 leads to impaired functions of lymphocytes in proliferation and
tumor rejection~\cite{scharffetter1998spontaneous,schmits1996lfa}.
Mutations in \textit{ITGB2} gene has been associated with type 1 leukocyte adhesion deficiency (LAD-1),
an autosomal-recessive inherited disease found in few families.
The disease is characterized by impair of lymphocytes in adherent-dependent functions, lack of accumulation
to the site of infection and recurrent bacterial and fungal infection~\cite{springer1987lymphocyte}.

Actin cytoskeleton is very important in T cell activation because it enhances activity of
LFA-1 by increasing its avidity and recruiting signaling molecules necessary for
downstream signaling~\cite{dustin2000immunological, van2000avidity}.
Cytoskeleton proteins binding to cytoplasmic domain of LFA-1 are thought to play an important role in
driving LFA-1 to aggregation on the cell surface, resulting in increased avidity.
Aggregation of LFA-1 has been demonstrated to be essential for lymphocytes to bind to the
ligand~\cite{van1994extracellular}.
Interestingly, \textit{RAC3}, \textit{PFN2}, and \textit{PPP1R12A}, which are involved in
actin cytoskeleton pathway (Figure~\ref{kegg_actin}), also expressed different ratios of alternative splice
forms between lines 6 and 7.
Some of these genes are also co-present in three other pathways
that involved in immune responses (Table~\ref{tab:integrin}).
This could be speculated that pre-mRNA splicing of these genes is co-regulated by splicing
regulators or some genetic factors.

\subsubsection*{Protein domains}
To understand the role of the alternative splice forms on the function of these genes,
transcripts were translated to protein sequences by ESTscan~\cite{iseli1999estscan}.
Protein sequences were then searched for annotated protein domains using
the standalone version of InterPro Scan~\cite{quevillon2005interproscan}.
Besides \textit{ITGB2}, other genes have alternative exons located in coding regions and
could potentially affect functional protein domains in some ways.
The exon with alternative 3$\prime$ splice site of \textit{RAC3} encode part of a protein
domain identified as small GTPase of Ras subfamily (ProSiteProfiles:PS5142 and SMART:SM00173).
Rac3 is highly homologous to Rac1 and has been reported to possess the ability of
promoting membrane ruffling, transformation, activation of
c-Jun transcriptional activity and a co-activator of NF$\kappa$B~\cite{,werbajh2000rac}.
Activated Rac also regulates production of superoxide in neutrophils and macrophages.

The alternative exon of \textit{PFN2} seems to disrupt the coding sequence that
encode profilin domain (Pfam:PF00235).
The profilin domain is essential for almost all organisms and its functions include
regulating actin polymerization, controlling complex network of molecular interaction
and transmitting signals from small-GTPase pathway.
It also binds to Rac effector molecules and a number of other ligands~\cite{witke2004role}.
The skipped exon of \textit{PPP1R12A} encodes part of a protein domain annotated as protein
phosphatase 1 (PIRSF:PIRSF038141).
The functional effects of alternative splicing on these functional domains have to be
further investigated by simulation or an experiment.

% Based on our gene model, the skipped exon of \textit{ITGB2} is located in the 5$\prime$ UTR;
% whereas, the skipped exon of \textit{HCK}, encodes protein tyrosine kinase (Pfam:PF07714).

\subsubsection*{Prediction of \textit{cis}-regulatory elements in alternative splicing exons}

Although alternative splicing of genes in all groups could be regulated by
\textit{cis-} and \textit{trans-}splicing factors,
genes in group II are more likely to be regulated by \textit{cis-}regulatory factors.
The ratios of isoform expression in this group were relatively stable within line,
but were significantly different between lines.
Investigation of nucleotide differences within exons of line 6 and 7 could reveal
a possible role of cSNPs in regulating alternative splicing in this group.
We obtained a sequence of an alternative exon from line 6 and used Human Splicing Finder (HSF)
to determine whether SNPs from line 7 could alter predicted ESEs or ESSs.
Results from some genes involved in cytokinesis are discussed in this section.
Exonic SNPs from line 6 and 7 are listed in Table~\ref{tab:deu_snps}.

>>>>

Marek's disease (MD) is an economically significant chicken disease that affects
the poultry industry worldwide with estimated annual loss of \$2 billion loss
has been reported from outbreaks~\cite{morrow2004marek}.  The disease is caused
by the highly oncogenic Marek's disease virus (MDV), an alphaherpesvirus that
induces T-cell lymphomas in susceptible birds.  Vaccination is the primary
control measure, which is effective in reducing incidence of tumor formation.
However, since MD vaccines are not sterilizing, they do not prevent infection or
horizontal spread of the virus.  As a consequence, MDV field strains that
overcome vaccinal protection have arisen repeatedly over time.  Therefore, there
is a need for sustainable alternative controls measures, such as improving
genetic resistance.

Many studies have reported strong associations between MHC alleles and
resistance or susceptibility to MD.  For example, chickens with MHC allele
B$^{21}$ are highly resistant in contrast to chickens with B$^{16}$ allele which
are highly susceptible.  ADOL lines 6 and 7, both share (B$^{2}$) allele, yet
exhibit different phenotypic responses to MD; e.g., challenge with JM/102W
strain with result in 0 and 100\% MD incidence for lines 6 and 7, respectively.
Thus, the major unanswered questions are what genetic factors, especially those
that are non-MHC, contribute to susceptibility and resistance of the disease and
what is the main contributing mechanism.

In the past decades, significant efforts have been spent to study variations in
a global gene expression between resistant and susceptible birds using
microarray and RNA-Seq methods in order to identify non-MHC genes that
contribute to resistance to MD~\cite{}.  However, none of the studies have
investigated differential expression of alternative isoforms, which are known to
play a significant role in many biological events including immune responses.
In addition, studies have shown that isoform expression levels can provide
better signatures for some diseases~\cite{zhang2013isoform}.  Changes of isoform
expression levels are governed partly by two types of {\em cis-}regulatory
elements: Exon Splicing Enhancer (ESE) and Exon Splicing Silencer (ESS), which
are located within an exon sequence.  A number of sequence motifs of ESE and ESS
have been identified in human and some other organisms and could be predicted
{\em in silico}.  Mutations that disrupt or create those motifs could alter
splicing patterns leading to aberrant alternative splicing.  A number of
disease-associated single-nucleotide polymorphsims in coding regions (cSNPs)
that affect ESEs and ESSs have been well characterized~\cite{blencowe2000exonic,
wang2007splicing}.
%In fact, 15\% of mutations that cause genetic diseases affect pre-mRNA splicing
%via this mechanism.
Therefore, variations in isoform expression could lead to identification of SNPs
that underlie genetic resistance to MD.  In this article, we reported
differential-expressed genes and isoforms that may contribute to resistance to
MD as well as cSNPs that can potentially affect isoform expression levels.

@@

To extensively study gene and isoform expression, we incorporated ENSEMBL gene
models release 73 with {\em de novo} and a reference-guided transcriptome
assembly to build custom gene models.  The models, therefore, include both
ENSEMBL annotated transcripts and putative transcripts and genes.  The advantage
of using custom gene models is it allows an investigation of unannotated genes
and isoforms, which is necessary for in-depth transcriptome study of an immune
system.

Many genes were found to be differentially expressed (DE) between control and
infected chickens in lines 6 and line 7.  While the number of unique
downregulated genes in both lines was approximately equal, the number of unique
upregulated genes in line 7 was much greater compared to line 6
(Figure~\ref{degenes_venn})~\cite{lee1983ontogeny}.  Interestingly, some genes
that were differentially expressed in both lines were regulated in the opposite
direction (Table~\ref{tab:opposite}).  Among genes downregulated in line 6 but
upregulated in line 7 were {\em LL} (lung lectin) and {\em SFTPA1}, which encode
a calcium-dependent C-type lectin and a lung surfactant protein respectively.
Both molecules are important in innate
immunity~\cite{hogenkamp2008chicken,kingma2006defense}.  {\em LIMS1} is involved
in cell differentiation and proliferation and {\em PPARG} is a suppressor of
{\em NF$\kappa$B}-mediated proinflamatory response.  On the other hand, nearly
all genes upregulated in line 6 but downregulated in line 7 are involved in cell
survival such as mRNAs splicing, cell growth, and protein synthesis, except CD7
whose function is involved in T cell-B cell interaction.  This difference
suggests that even at this stage of infection in line 6, the lytic phase could
be subsided, therefore, only genes important for survival of cell are
upregulated in line 6.  In comparison, the lytic phase in line 7 may still
continue and as a result, genes involved in immune responses are still
upregulated.  In addition, type I interferon ({\em IFN-$\gamma$} and {\em
IFN-$\beta$}) as well as {\em INF-$\alpha$3} were found to be highly upregulated
in infected chickens in both lines 6 and 7 (Table~\ref{tab:cytokines}).
However, expressions of genes encoding their corresponding receptors were not
different in line 6, but upregulated in line 7.  This could also reflect the
ongoing immune responses in line 7.

Some DE genes have been reported from previous microarray studies were also
found to be differentially expressed in this study.  For example, B6.1 (Bu-1)
gene is known to be down-regulated approximately 2.27 fold in susceptible
chicken with the MHC allele B\textsuperscript{19} at 4
d.p.i~\cite{sarson2008transcriptional}.  It was also found to be down-regulated
about 3-fold in line 7 (susceptible).  Similarly, {\em GMZA} gene reported to be
upregulated across genetically different chickens (B$^{19}$, B$^{21}$ alleles),
was also found to be highly upregulated.  In contrast, some genes that have been
reported to be highly expressed in resistant chickens were downregulated in both
lines. Those genes are {\em AMIGO2, MMP13} and {\em CLEC3B}, which were found to
be downregulated more than 2-fold~\cite{sarson2008transcriptional}.
% Several genes involved in innate immune response were differentially expressed
% in resistant chickens: \textit{DDT, NMU, VIPR2} and \textit{HSP5}. Of all
% those genes, only HSP5 was upregulated.
Other immune genes reported to be highly expressed in susceptible chickens (line
7$_{2}$) including {\em AVD, ART1, NOS2, CXCL13L2, MX1} and {\em
SOCS1}~\cite{smith2011systems} were also found to be highly upregulated in both
line 6 and 7.  However, our results show the similar expression patterns for
{\em IL6} and {\em IL18}, which were only upregulated in line 7 in early stage
of infection (3-5 d.p.i).

In contrast, {\em IL15} has been reported non differential-expressed between
control and infected chickens in line 6$_{1}$ and
7$_{2}$~\cite{kaiser2003differential}; however, it was only upregulated in line
7 in our study.  Expression of {\em IL15} is induced by {\em TLR9}, which binds
to non-methylated CpG residues present in genomes of many DNA viruses, including
herpes simplex virus.  This cytokine auto-regulates the expression of {\em
CD40}, which is a transmembrane receptor required for activation of macrophages
by CD4 T cells.  Consequently, {\em CD40} was only upregulated in line 7 (data
not shown).

@@

To determine pathways that were perturbed during the infection, data were
analyzed by GOSeq, which accounted for gene lengths bias unique for RNA-Seq
data~\cite{young2010method}.  Significantly perturbed pathways (FDR $< 0.1$)
from both lines that involved in immune response include TLR signaling pathway,
cytokine-cytokine receptor interaction, intestinal immune network for IgA
production, and cell adhesion molecules (CAMs)~\ref{line67_kegg}.  Some other
pathways important in response to viral infection and only significantly
enriched in line 7 include phagosome, apoptosis, RIG-I-like receptor signaling
pathway, NOD-like receptor signaling pathway, and lysosome.
Figure~\ref{kegg_phagosome} is a pictorial example of differentially expressed
genes in the phagosome pathway.  Although MHC I gene ({\em BF1}) was
differentially expressed in both lines, other genes involved in expressing newly
synthesized MHC I were only upregulated in line 7 suggesting that new MHC I
molecules were actively produced and expressed on cell surface.
% (the minor class I is upregulated but the major class I is downregulated)
Furthermore, gene ontology analysis of biological processes (GO:BP) shows that
categories involved in both adaptive and innate immune responses were enriched
in line 7 (supplementary material).  On the other hand, only categories involved
in innate immune responses were enriched in line 6.  In addition, enrichment of
apoptosis pathway in line 7 indicates that the programmed cell death could be
induced by CTL response to eliminate the viruses.

@@

Due to those issues, it is not feasible to accurately estimate expression of
isoforms, especially when gene annotation is constructed from {\em de novo}
assembly~\cite{trapnell2013differential}.  To avoid these issues, we chose to
study exon expression instead of isoform expression.  Using MISO with
exon-centric method, only reads spanning across a few exons are used and only
exons involved in a splicing event are examined.  The expression of exon
inclusion is calculated as Percent Spliced In (Psi or $\Psi$), which can be used
to infer the portion of transcripts that include the exon in each
sample~\cite{Katz:2010iv}.  In this study, we investigated the three most common
alternative splicing events in vertebrates, which are skipped exons (SE), an
alternative $3\prime$ (A3SS) and $5\prime$ (A5SS) splice site.

@@

Group I (Table~\ref{tab:line67i_diff_line67u_one}) includes genes with $\Psi$s
that were up- or down-regulated in infected chickens in line 6 only.  This group
includes {\em BCL11B} (B-cell CLL/lymphoma 11B zinc finger proteins), a B-cell
lymphoma associated C2H2-type zinc finger protein encoding gene, which functions
as a tumour-suppressor in T-cell lymphoma in human.  According to homologous
alignments on UCSC genome browser, a splice form with the skipped exon is
similar to mouse {\em BCL11B isoform b}.  The skipped exon was expressed 30\% in
infected line 6 chickens; whereas it was rarely expressed (4-7\%) in the control
line 6 and both groups in line 7.  The skipped exon was not found to encode any
known protein domain, however, it is in the middle of two adjacent C2H2-type
finger protein domains.  Another gene related to B and T cell lymphomas is {\em
SIK2} (salt-inducible kinase 2).  This gene has been reported to have a negative
effect on T cell lymphomas by limiting the transcription of HTLV-1
virus~\cite{tang2013lkb1}.  Some other genes are important in pre-mRNA splicing
including {\em TRA2A}, {\em SRSF6} and {\em GEMIN6}.  {\em TRA2A} (transformer-2
protein homolog alpha) encodes RNA recognition motif (RRM), the skipped exon is
not found to encode any known protein domain.  {\em SRSF6} (SR splicing factor
6) encodes a nuclear protein that belongs to the splicing factor protein family.
{\em GEMIN6} plays a role in the assembly of spliceosomal snRNP in cytoplasm.
These genes could play a significant role in regulating inclusion of alternative
exons.  A gene in this group that might be important for innate immune responses
is {\em RAC3} (Ras-related C3 botulinum toxin substrate 3).  This gene encodes
small GTPases, belonging to Ras family, that regulates a wide variety of
cellular events including cell growth, cytoskeletal reorganization, and the
activation of protein kinases.  The role of small GTPases in immune responses is
discussed further below.

In group II (Table~\ref{tab:line67i_diff_line67u_two}), $\Psi$ values were
relatively stable in control and infected chickens within line, but not between
lines.  Genes that could play an important role in immune responses are {\em
ITGB2} and {\em HCK}.  {\em HCK} transmits signal from cell surface receptors
such as {\em FCGR1A, FCGR2A, IL2, IL6, IL18}, and integrins ({\em ITGB1,
ITGB2}).  {\em ITGB2} (CD18) encodes subunit $\beta_{2}$ integrin of {\em LFA-1}
and {\em CR3} receptors.  {\em LFA-1} plays an important role in adhesion of
lymphocytes with other cells.  {\em CR3} binds to a vast array of ligands and
molecules including complement C3bi, microbial proteins, ICAM-1 and -2, ECM
proteins, and coagulation proteins~\cite{}.  It plays a significant role in
neutrophils and monocytes activation including phagocytosis, adhesion and
migration~\cite{}.  The role of {\em ITGB2} in immune responses is discussed
further in the next section.  {\em DYNLT1,DYNLL2, SEPT11, PFN2}, and {\em
ZDHHC7} are also involved in cell rearrangement and cytokinesis.  In particular,
{\em DYNLT1} and {\em DYNLL2} are dynein proteins that have been demonstrated to
regulate T cell activation by driving T cell receptor microclusters (TCR-MCs)
toward the center of immune synapse~\cite{hashimoto2011dynein}.

Group III (Table~\ref{tab:line67i_diff_line67u_three}) includes genes that
exhibit differential isoform expression only in infected line 7.  A number of
genes in this group encode proteins that are parts of spliceosome: {\em SRSF3},
{\em HNRNPDL}, {\em SFSWAP}, {\em THOC1}, {\em RNPC3} and {\em SRSF5}.  Two
genes involved in cell-cell contact regulated by integrins are also in this
group.  {\em PPP1R12A} is a myosin phosphatase that regulates the interaction of
actin and myosin downstream of GTPase Rho proteins.  {\em PODXL} encodes
PODX-like protein that functions in integrin-dependent manner as both
pro-adhesive and anti-adhesive molecules.  This protein is involved in
cell-to-cell contact, cell trafficking, and cancer progression~\cite{}.

The last group (Group IV, Table~\ref{tab:line67i_diff_line67u_three}) only has
one gene, {\em GOSR1}.  The $\Psi$ value of this gene was less than 0.20 cutoff
in control and infected chickens in line 6, but it is significantly different
between infected chickens in line 6 and 7.  Also, there is a significant
difference between control and infected chickens in line 7.  This gene encodes a
trafficking membrane protein important for transporting proteins from {\em cis-}
to {\em trans-}golgi network.

\subsubsection*{Roles of {\em LFA-1} and actin cytoskeleton pathway in T cells activation}

Integrins are a special kind of receptors that transmit signals bidirectionally
across the cell membrane.  They are heterodimeric composed of an $\alpha$
(large) and a $\beta$ (small) subunits~\cite{wang2010immunopathologies}.
$\beta_{2}$ (CD18) subunit encoded by {\em ITGB2} is expressed on lymphocytes
and antigen presenting cells (APCs) as a component of {\em LFA-1} and {\em CR3}
receptors.  LFA-1 binds to its ligand ICAM-1 to help form a synapse that bring
APC and T cell together to initiate antigen presentation leading to T cell
activation~\cite{dustin2000immunological}.  Absence of LFA-1 leads to impaired
functions of lymphocytes in proliferation and tumor
rejection~\cite{scharffetter1998spontaneous,schmits1996lfa}.  Mutations in {\em
ITGB2} gene has been associated with type 1 leukocyte adhesion deficiency
(LAD-1), an autosomal-recessive inherited disease found in few families.  The
disease is characterized by impair of lymphocytes in adherent-dependent
functions, lack of accumulation to the site of infection and recurrent bacterial
and fungal infection~\cite{springer1987lymphocyte}.

Actin cytoskeleton is very important in T cell activation because it enhances
activity of LFA-1 by increasing its avidity and recruiting signaling molecules
necessary for downstream signaling~\cite{dustin2000immunological,
van2000avidity}.  Cytoskeleton proteins binding to cytoplasmic domain of LFA-1
are thought to play an important role in driving LFA-1 to aggregation on the
cell surface, resulting in increased avidity.  Aggregation of LFA-1 has been
demonstrated to be essential for lymphocytes to bind to the
ligand~\cite{van1994extracellular}.  Interestingly, {\em RAC3, PFN2,} and {\em
PPP1R12A}, which are involved in actin cytoskeleton pathway
(Figure~\ref{kegg_actin}), also expressed different ratios of alternative splice
forms between lines 6 and 7.  Some of these genes are also co-present in three
other pathways that involved in immune responses (Table~\ref{tab:integrin}).
This could be speculated that pre-mRNA splicing of these genes is co-regulated
by splicing regulators or some genetic factors.

\subsubsection*{Protein domains}

To understand the role of the alternative splice forms on the function of these
genes, transcripts were translated to protein sequences by
ESTscan~\cite{iseli1999estscan}.  Protein sequences were then searched for
annotated protein domains using the standalone version of InterPro
Scan~\cite{quevillon2005interproscan}.  Besides {\em ITGB2}, other genes have
alternative exons located in coding regions and could potentially affect
functional protein domains in some ways.  The exon with alternative 3$\prime$
splice site of {\em RAC3} encode part of a protein domain identified as small
GTPase of Ras subfamily (ProSiteProfiles:PS5142 and SMART:SM00173).  Rac3 is
highly homologous to Rac1 and has been reported to possess the ability of
promoting membrane ruffling, transformation, activation of c-Jun transcriptional
activity and a co-activator of NF$\kappa$B~\cite{,werbajh2000rac}.  Activated
Rac also regulates production of superoxide in neutrophils and macrophages.

The alternative exon of {\em PFN2} seems to disrupt the coding sequence that
encode profilin domain (Pfam:PF00235).  The profilin domain is essential for
almost all organisms and its functions include regulating actin polymerization,
controlling complex network of molecular interaction and transmitting signals
from small-GTPase pathway.  It also binds to Rac effector molecules and a number
of other ligands~\cite{witke2004role}.  The skipped exon of {\em PPP1R12A}
encodes part of a protein domain annotated as protein phosphatase 1
(PIRSF:PIRSF038141).  The functional effects of alternative splicing on these
functional domains have to be further investigated by simulation or an
experiment.

% Based on our gene model, the skipped exon of \textit{ITGB2} is located in the
% 5$\prime$ UTR; whereas, the skipped exon of \textit{HCK}, encodes protein
% tyrosine kinase (Pfam:PF07714).

\subsubsection*{Prediction of {\em cis}-regulatory elements in alternative splicing exons}

Although alternative splicing of genes in all groups could be regulated by {\em
cis-} and {\em trans-}splicing factors, genes in group II are more likely to be
regulated by {\em cis-}regulatory factors.  The ratios of isoform expression in
this group were relatively stable within line, but were significantly different
between lines.  Investigation of nucleotide differences within exons of line 6
and 7 could reveal a possible role of cSNPs in regulating alternative splicing
in this group.  We obtained a sequence of an alternative exon from line 6 and
used Human Splicing Finder (HSF) to determine whether SNPs from line 7 could
alter predicted ESEs or ESSs.  Results from some genes involved in cytokinesis
are discussed in this section.  Exonic SNPs from line 6 and 7 are listed in
Table~\ref{tab:deu_snps}.
